Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-07-17 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
Last Revision: small changes to improve readibility

\begin{abstract} 
%CAUSE --> EFFECT --> MOTIVATION (P1)
  A critical aspect of multiphase flow in porous media is the displacement efficiency that measures the amount of fluid that can be pushed by another fluid driven by pressure gradient. Migration of contaminants and reservoir waterflooding are typical applications where understanding the dynamics of immiscible fluid displacement helps mitigating water resources contamination and improving hydrocarbons production, respectively. Due to large viscosity ratios, flow instabilities at fluids' interface may arise leading to the formation of fingers therefore creating an uneven front with elongation at the outside edge of fluids interface with strong impact on the displacement efficiency.  
%CAUSE --> EFFECT --> GAPS IN KNOWLEDGE --> AIMS OF THE WORK (P2)
Initial studies of viscous instabilities in Hele-Shaw cells indicated that the development of fingers mostly depends on mobility and capillary forces, however heterogeneity of the porous domain may also affect the onset of instabilities. Therefore, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. 
%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
The model used here is based on a novel control volume finite element method (CVFEM) formulation with families of FE-pairs, \PN[n]{m} and  \PNDG[n]{m}, specially tailored for Darcean flows. Dynamic mesh adaptivity enables capturing fingers development whilst saving computational overheads. Numerical experiments were performed to investigate the impact of viscosity ratio and heterogeneity on Saffmann-Taylor instabilities.
%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
Numerical simulations demonstrated that the heterogeneity of the domain triggers the early-onset formation of fingers under prescribed viscosity ratio conditions. Also, effective numerical capture of growth (in particular tip-splitting) and coalescence of dendritic finger branching induced by large viscosity ratio largely depends on mesh resolution at the fluids interface.

@@

\section{Introduction}\label{section:intro}
Numerical investigation of multiphase flows in porous media has attracted the attention of the scientific community over the past 40 years. Characterisation and prediction of such flows serve as the foundation of hydrocarbon reservoir and groundwater studies.% \citep{white_1981}. %Underground coal gasification is another important field of interest and, more recently, due to the role of green house gases (GHG) emissions in the global climate change, several research work have focused on CO$_{2}$ migration and trapping mechanisms in carbon capture utilisation and storage (CCUS) operations \citep{spycher_2003, self_2012, jiang_2011}.

Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. %Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities,][]{saffman_1959}. 
This work focused on flow instabilities in two-phase systems due to viscous and stress forces, often referred as viscous instabilities or viscous fingering.

%Viscous flow instabilities can be found across several disciplines and scales, from chemical separation processes to geological reservoir fluids. 
\citet{muskat_1934} used Hele-Shaw cells (\ie parallel flat plates separated by an infinitesimal gap) to study fluid flow and the impact on the capillary number $\left(\text{N}_{c}\right)$ in the flow dynamics.
%investigated fluid flow in Hele-Shaw cells, \ie parallel flat plates separated by an infinitesimal gap, and the impact on the capillary number ($N_{c}$),
%\begin{equation}
%N_{c} = \frac{\mu U}{\gamma}.\label{eq:capillary_number}
%\end{equation}
 %In Eqn.~\ref{eq:capillary_number}, $\gamma$ is the surface tension and $U$ is the characteristic velocity of the moving interface. 
This experimental apparatus enabled instabilities to be qualitatively investigated by simplifying the flow (in both porous and non-porous media) into a 2D problem. \citet{mclean_1981} developed a semi-analytic solution for flows in Hele-Shaw cells which was later used by \citet{guan_2003} to investigate fingers' formation, dimensions and branchiness.

@@

\medskip 
Multi-fluid flow dynamics in porous media are described by continuity and momentum conservative equations for each fluid (or phases) with coupling mass terms (\ie density and saturation) appearing in both sets of equations. Finite difference methods (FDM) have been extensively used in most industry-standard reservoir simulators however, they are often limited to relatively simple geometries and often lead to excessive numerical dispersion when strong heterogeneity is present~\citep{chavent_1986}.

The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient than FDM to solve fluid flow dynamics in complex geometries. Among FEM-based formulations for porous media flows, the control volume finite element methods~\citep[CVFEM,][]{fung_1992} was designed to guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In traditional CVFEM formulations, pressure and velocity are interpolated using piecewise linear FE basis functions, while material properties and flow conditions (\eg phase saturation, density, temperature, species concentration etc) are interpolated with CV basis functions~\citep{voller_2009}.

>>>>

\begin{abstract} 
%CAUSE --> EFFECT --> MOTIVATION (P1)
  A critical aspect of multiphase flow in porous media is the displacement efficiency, which measures the amount of fluid that can be pushed by another fluid driven by pressure gradient. Migration of contaminants and reservoir waterflooding are typical applications where understanding the dynamics of immiscible fluid displacement helps mitigating water resources contamination and improving hydrocarbons production, respectively. Due to large viscosity ratios, flow instabilities at fluids' interface may arise leading to the formation of fingers, \ie uneven fronts with elongation at the outside edge of fluids interface with strong impact on the displacement efficiency.  
%CAUSE --> EFFECT --> GAPS IN KNOWLEDGE --> AIMS OF THE WORK (P2)
Initial studies of viscous instabilities indicated that the development of fingers mostly depends on mobility and capillary forces, however heterogeneity of the porous domain may also affect the onset of instabilities. Therefore, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. 
%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
The model used here is based on a novel control volume finite element method (CVFEM) formulation with families of FE-pairs, \PN[n]{m} and  \PNDG[n]{m}, specially tailored for Darcean flows. Dynamic mesh adaptivity enables capturing fingers development whilst saving computational overheads. Numerical experiments were performed to investigate the impact of viscosity ratio and heterogeneity on Saffmann-Taylor instabilities.
%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
Numerical simulations demonstrated that the heterogeneity of the domain triggers the early-onset formation of fingers under prescribed viscosity ratio conditions. Also, effective numerical capture of growth (in particular tip-splitting) and coalescence of dendritic finger branching induced by large viscosity ratio largely depends on mesh resolution at the fluids interface.

@@

\section{Introduction}\label{section:intro}
Numerical investigation of multiphase Darcy flows has attracted the attention of the porous media community over the past 40 years. Characterisation and prediction of such flows are the foundation of groundwater and oil $\&$ gas reservoirs studies.% \citep{white_1981}. %Underground coal gasification is another important field of interest and, more recently, due to the role of green house gases (GHG) emissions in the global climate change, several research work have focused on CO$_{2}$ migration and trapping mechanisms in carbon capture utilisation and storage (CCUS) operations \citep{spycher_2003, self_2012, jiang_2011}.

Description of physics and mechanisms of multiphase porous media flows was reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. %Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities,][]{saffman_1959}. 
This work focused on flow instabilities in two-phase systems due to viscous and stress forces, often referred as viscous instabilities or viscous fingering.

%Viscous flow instabilities can be found across several disciplines and scales, from chemical separation processes to geological reservoir fluids. 
\citet{muskat_1934} used Hele-Shaw cells (\ie parallel flat plates separated by an infinitesimal gap) to study fluid flow and the impact of the capillary number $\left(\text{N}_{c}\right)$ on the flow dynamics.
%investigated fluid flow in Hele-Shaw cells, \ie parallel flat plates separated by an infinitesimal gap, and the impact on the capillary number ($N_{c}$),
%\begin{equation}
%N_{c} = \frac{\mu U}{\gamma}.\label{eq:capillary_number}
%\end{equation}
 %In Eqn.~\ref{eq:capillary_number}, $\gamma$ is the surface tension and $U$ is the characteristic velocity of the moving interface. 
This experimental apparatus enabled instabilities to be qualitatively investigated by simplifying the flow (in both porous and non-porous media) to a 2D problem. \citet{mclean_1981} developed a semi-analytic solution for flows in Hele-Shaw cells which was later used by \citet{guan_2003} to investigate fingers' formation, dimensions and branchiness.

@@

\medskip 
Multi-fluid flow dynamics in porous media are described by continuity and momentum conservative equations for each fluid with coupling terms appearing in both sets of equations. Finite difference methods (FDM) have been extensively used in most industry-standard reservoir simulators to solve this set of equations, however they are often limited to relatively simple geometries and often lead to excessive numerical dispersion when strong heterogeneity is present~\citep{chavent_1986}.

The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient than FDM to solve fluid flow dynamics in complex geometries. Among FEM-based formulations for porous media flows, the control volume finite element method~\citep[CVFEM,][]{fung_1992} was designed to guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In traditional CVFEM formulations, pressure and velocity are interpolated using piecewise linear FE basis functions, while material properties and flow conditions (\eg phase saturation, density, temperature, species concentration etc) are interpolated with CV basis functions~\citep{voller_2009}.
